Group polarization replications may be marred by high false discovery rates

### 2024-12-28 GroupPolarizationStatmod/paper/ms.tex
calculating quantiles and writing analysis

@@

\subsection{How study design can induce false detections of group polarization}

@@

\subsection{Outcomes: Estimating and limiting false detection rates}

To calculate the false detection rate we begin by simulating 
pre- and post-deliberation 
opinions by drawing $N_e$ samples from normal latent opinion
distributions that generated spurious group polarization, i.e., with
parameters labelled $\beta_e$ found in the hillclimbing step. Binning these
random values simulates observations. $N_e$ is the number
of participants in experimental condition $e$. Each trial $i$ of $N_e$
simulated pre- and post-deliberation observations is then inspected to see
if spurious group polarization occurred in the simulation. Spurious group
polarization is determined by first inferring the latent parameters of each
simulated experimental trial by fitting an ordered probit model. 
The inferred parameters are then used to calculate the effect size for that
trial, Cohen's $d$, defined as~\cite[p. 331]{Liddell2018}
\begin{equation}
  d = \frac{\mupost - \mupre}{\sqrt{\frac{1}{2}(\sigmapost^2 + \sigmapre^2)}}.
\end{equation}
\noindent
The family-wise error rate is, $\alpha = \Pr(D|\neg E)$, is the 
fraction of trials with $d > d^*$. To calculate the family-wise error rate
for a study, $s$ which contains several experimental conditions, $e$,
\begin{equation}
  \alpha_s = \frac{1}{|s|} \sum_{e \in s} \alpha_e
\end{equation}
\noindent
The grand mean across all studies for the corpus is 
\begin{equation}
  \alpha_\mathrm{all} = \frac{1}{|E|} \sum_{e \in E} \alpha_e.
\end{equation}
With $\alpha$ calculated, we calculate the false discovery rate using
Equation~\ref{eq:fdr}.

@@

\section{Results}\label{results}

\subsection{Ninety percent of detections identified as plausibly false}

We identified 54 of 60 experimental 
conditions where a $N \to \infty$ model falsely indicated group polarization 
instead of simple consensus.

\subsection{False detection rates exceed ??\% even at ``high significance''}

@@

Our analysis brings some additional questions into focus. Most
significantly, recall that even the commensurate ordered probit Bayesian
model often failed to accurately detect negative results. A more rigorous
analysis would include additional sources of variance that will likely
further inflate the family-wise error rate, $\alpha$, and the false
discovery rate~\cite{Yarkoni2022}. First and foremost, there is variance
within each experimental deliberation group; no group polarization study
accounts for this, instead treating participants in each experimental
condition as if they were one large group. Furthermore, survey data are
known to be noisy~\cite{Zaller1992}. People report opinions differently
over time for no apparent reason (i.e., opinions are \emph{unstable}).
Context matters: the order in which a survey question is asked, and which
question framing is chosen among logically equivalent alternatives, can
both significantly influence participant responses. The effect of 
accounting for these sources of variance must be understood to estimate
$\alpha$ (and power, $W$) for the design of the next generation of group
polarization experiments.

>>>>

If an extremist opinion falls and it is measured with a Likert scale, will it make a noise?

@@

\subsection{How study design can induce false detections of group polarization}
\label{subsec:false-detections}

@@

\subsection{Estimating and limiting false detection rates}

To calculate the false detection rate we begin by simulating 
pre- and post-deliberation 
opinions by drawing $N_e$ samples from normal latent opinion
distributions that generated spurious group polarization, i.e., with
parameters labelled $\beta_e$ found in the hillclimbing step. Binning these
random values simulates observations. $N_e$ is the number
of participants in experimental condition $e$. Each trial $i$ of $N_e$
simulated pre- and post-deliberation observations is then inspected to see
if spurious group polarization occurred in the simulation. Spurious group
polarization is determined by first inferring the latent parameters of each
simulated experimental trial by fitting an ordered probit model~\cite{Liddell2018}. 
The inferred parameters are then used to calculate the effect size for that
trial, Cohen's $d$, defined as~\cite[p. 331]{Liddell2018}
\begin{equation}
  d = \frac{\mupost - \mupre}{\sqrt{\frac{1}{2}(\sigmapost^2 + \sigmapre^2)}}.
  \label{eq:cohens}
\end{equation}
\noindent
The family-wise error rate is, $\alpha_{e} = \Pr(D|\neg E)$, is the 
fraction of trials, 
with $d_{ei} > d^*$ for the $i^\text{th}$ trial for experimental
condition $e$, written
\begin{equation}
  \alpha_e = \frac{1}{N_T} \sum_i^{N_T} \Theta_i,
\end{equation}
\noindent
where
\[
  \Theta_i = 
    \begin{cases}
      1 & \text{ if } d_{ei} > d^* \\
      0 & \text{ otherwise.}
    \end{cases}
\]
and $N_T = 1000$ is the number of trials.

To calculate the family-wise error rate
for a study, $s$ which contains several experimental conditions, $e$,
\begin{equation}
  \alpha_s = \frac{1}{|s|} \sum_{e \in s} \alpha_e
\end{equation}
\noindent
The grand mean across all studies for the corpus is 
\begin{equation}
  \alpha_\mathrm{all} = \frac{1}{|E|} \sum_{e \in E} \alpha_e.
\end{equation}
With $\alpha$ calculated, we calculate the false discovery rate using
Equation~\ref{eq:fdr}.

@@

\section{Analysis}\label{analysis}

Our analysis estimates how the probability that a published detection of group
polarization is actually a false detection. We first identified latent opinion
parameters that yield spurious group polarization in an asymptotic $N\to\infty$
model. This gives us our first suggestion that experimental conditions may be
plausibly false. Briefly, we identified latent opinion distribution parameters
that generated spurious group polarization in 54 out of 57 experimental
conditions. In other words, 95\% of all group polarization detections are
plausibly spurious due to ceiling artifacts from the continuous latent
opinions-ordinal measurement experimental design. We then analyze family-wise
error and false detection rates for each experimental condition and aggregated 
up to the study level and across all studies and experimental conditions. 
Family-wise error rates varied from a minimum of 0.097 for the
\texttt{Feminists-Experimental} condition from Myers (1975)~\cite{Myers1975a}
and a maximum of 0.67 for the \texttt{COSprings-CivilUnions} condition from
Schkade, Sunstein, and Hastie (2010)~\cite{Schkade2010}. The median family-wise error
rate is 0.28 for Krizan and Baron's experimental condition
\texttt{NoOutgroupScenario1}. We close by inverting our procedure of calculating
family-wise error rates to calculating what $d^*$ would be required in each of the 54
experimental conditions to reduce the family-wise error rate to $\alpha = 0.05$.

\subsection{Nearly all group polarization detections are plausibly false}

We used our formal distribution model of group polarization experiments
to identify plausible latent simple consensus distributions that generate spurious
group polarization in observed ordinal distributions 
in the asymptotic case of taking $N \to \infty$ ordinal measurements.
Across the ten studies we analyzed, we excluded 3 out of 60 experimental
conditions that did not claim to be positive findings of group polarization.
Of the remaining 57, we identified plausible $\beta$ that generate spurious
group polarization in 54 experimental conditions. In the other three, we did
identify $\beta$ that generate spurious group polarization, but the initial
simulated distributions were highly polarized. Two of these
simulations of experimental conditions from Schkade, et al., (2010) because
the histogram plots of the observed opinion distributions are not
polarized~\cite[Figure 1, p. 234]{Schkade2010}. The other condition where we
rejected candidate $\beta$ was in a question about a ``bad teacher'' in Myers
(1975)~\cite{Myers1975a}, where lower values signify increasingly 
worse impressions of the teacher and higher values signify increasingly good
impressions. It is unlikely that any participants would rate a ``bad teacher''
as good, so we excluded this from our analysis as well. This exclusion may be
overly generous since Myers (1975) did not provide details beyond mean values
and opinion shifts, and the article was published before the importance
of open data best practices was recognized.

\clearpage

\begin{table}[ht]
\rotatebox{90} {
  \input{tex/quantile_table.tex}
}
\end{table}

\subsection{Half of experiments exceed a 75\% false discovery rate at ``medium'' significance}

The $N\to\infty$ model enabled us to identify pre- and post-deliberation latent
opinion distribution parameters that caused spurious group polarization, denoted
$\beta_e$ for experimental condition $e$. Using these $\beta$ we drew 
$N_\mathrm{emp}$ samples from the pre- and post-deliberation latent opinion 
distributions, then binned them (Section~\ref{subsec:false-detections}) 
to simulate participants reporting their pre- and post-deliberation opinions. 
To estimate optimistic family-wise error and false detection rates, we fit an
ordered probit model to the simulated observations and obtained estimates of
the group polarization effect size, $d$ (Equation~\ref{eq:cohens}).
If the effect size was greater than a significance value, $d > d^*$, then there
was a detection $D$ of group polarization, even though by design the simulated latent mean
group opinion was constant, i.e., there in fact was no effect, written $\neg E$.
The family-wise error rate is the calculation of $\Pr(D|\neg E)$ as the frequency with which
$d > d^*$ out of 1000 simulation trials. The false detection rate is the family-wise
error rate scaled using the base rate of effects and the statistical power
(Equation~\ref{eq:fdr}). We test three difference significance values,
$d = 0.2,~0.5$ and $0.8$, corresponding to what Cohen prescribed to use for
``low'', ``medium'', and ``high'' significance, respectively.

@@

Our analysis brings some additional questions into focus. Most significantly, recall that even
the more appropriate ordered probit Bayesian model often failed to accurately detect negative
results with the number of participants used in the original papers. Many of the studies
analyzed here did not use any statistical test to establish significance; those that did used
some form of a $t$-test, and it is not clear if the $t$-test that was used 
accounted for a change in pre- and post-deliberation opinion variance, which we expect due to
consensus.

The ordered probit model we used still likely underestimates the variance in outcomes, which
could require even larger sample sizes. A more rigorous analysis
would include additional sources of variance that will likely further inflate the Type I
error rate, $\alpha$, and the false discovery rate~\cite{Yarkoni2022}. First and foremost,
there is variance within each experimental deliberation group; no group polarization study
accounts for this, instead treating participants in each experimental condition as if they were
one large group. Furthermore, survey data are known to be noisy~\cite{Zaller1992}. People
report opinions differently over time for no apparent reason (i.e., opinions are
\emph{unstable}).  Context matters: the order in which a survey question is asked, and which
question framing is chosen among logically equivalent alternatives, can both significantly
influence participant responses. The effect of accounting for these sources of variance must be
understood to estimate $\alpha$ (and power, $W$) for the design of the next generation of group
polarization experiments.
